---
title: "scratchpad"
author: "Ethan Horn, David Lopez, Alec Libroia"
date: "2023-11-17"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidymodels)
library(lubridate)
library(gghighlight)
library(ggrepel)
set.seed(19)
```

# Data Preprocessing

Load in each dataset, store them in the `raw` object, to reduce namespace pollution. The raw datasets are not needed past the data preprocessing step, so they should not heavily pollute the namespace.

```{r load_data, message=FALSE}
raw <- {}
raw$hri.emer           <- jsonlite::fromJSON('datasets/hri_emergencyvisits.json')
raw$hri.hosp           <- jsonlite::fromJSON('datasets/hri_hospitalization.json')
raw$hri.mort           <- jsonlite::fromJSON('datasets/hri_mortality.json')

raw$co2                <- read_csv('datasets/Atmospheric_CO2_Concentrations.csv')

raw$temp.state         <- read_csv('datasets/climdiv_state_year.csv')
raw$temp.state.states  <- read_csv('datasets/model_state.csv')
raw$temp.state.regions <- read_csv('datasets/state-region.csv')
```

Now we must tidy the datasets, giving each column a consistent naming convention and logical type, and removing excess columns.

```{r tidy_data}
hri.emer <- raw$hri.emer$tableResult %>%
  filter(hatchingId == -1) %>%  # Filter out "Unstable" points of data
  select(geo, geoId, year, dataValue) %>%
  rename(state           = geo,
         state.id        = geoId,
         emer.per.capita = dataValue) %>%
  mutate(state.id        = as.numeric(state.id),
         year            = as.numeric(year),
         emer.per.capita = as.numeric(emer.per.capita))

hri.hosp <- raw$hri.hosp$tableResult %>%
  filter(hatchingId == -1) %>%  # Filter out "Unstable" points of data
  select(geo, geoId, year, dataValue) %>%
  rename(state           = geo,
         state.id        = geoId,
         hosp.per.capita = dataValue) %>%
  mutate(state.id        = as.numeric(state.id),
         year            = as.numeric(year),
         hosp.per.capita = as.numeric(hosp.per.capita))

hri.mort <- raw$hri.mort$tableResult %>%
  filter(suppressionFlag == 0) %>% # Filter out "Suppressed" points of data
  select(geo, geoId, year, dataValue) %>%
  rename(state             = geo,
         state.id          = geoId,
         mortalities.total = dataValue) %>%
  mutate(state.id          = as.numeric(state.id),
         year              = as.numeric(year),
         mortalities.total = as.numeric(mortalities.total))

co2 <- raw$co2 %>%
  filter(Unit == "Parts Per Million") %>%
  select(Date, Value) %>%
  rename(date = Date,
         ppm  = Value) %>%
  mutate(date  = str_replace(date, "M", "-"),
         date  = paste0(date, '-01'),
         date  = as.Date(date),
         year  = lubridate::year(date),
         month = lubridate::month(date))

temp.state <- raw$temp.state %>%
  filter(year >= 2000) %>%
  left_join(raw$temp.state.states,  by = join_by(fips)) %>%
  left_join(raw$temp.state.regions, by = join_by(STATE_NAME)) %>%
  select(year, temp, STUSAB, STATE_NAME, region) %>%
  rename(state.code   = STUSAB,
         state        = STATE_NAME,
         state.region = region)
```

# Exploratory Data Analysis

The following sections visualize each dataset, to analyze for abnormalities in each dataset.

## Emergency Room Visits

The first graph plots each observation in the emergency room visits dataset, by the number of visits per capita and the year the observation was recorded. Each point on the plot represents a different state's record. Also plotted in this graph is the least squares regression line, and an imposed line graph that tracks the average visits per capita, nationwide. The second graph shows the variability between years in this dataset. The third graph highlights the outliers in the second graph.

```{r emer_cor}
ggplot(hri.emer, aes(year, emer.per.capita)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  geom_line(aes(year, emer.per.capita), data = aggregate(emer.per.capita ~ year, hri.emer, mean)) +
  labs(title = "Yearly HRI Emergency Room Visits per Capita by State",
       x     = "Year",
       y     = "Emergency Room visits per Capita (100,000)")

ggplot(hri.emer, aes(factor(year), emer.per.capita)) +
        geom_boxplot() +
        labs(title = "Yearly HRI Emergency Room Visits per Capita by State",
             x     = "Year",
             y     = "Emergency Room visits per Capita (100,000)") +
  scale_x_discrete(guide = guide_axis(angle = 90))

ggplot(hri.emer, aes(factor(year), emer.per.capita)) +
  geom_boxplot(outlier.colour = "red") +
  geom_text_repel(data = hri.emer %>% group_by(year) %>% filter(emer.per.capita > quantile(emer.per.capita, 0.75) + 1.5*IQR(emer.per.capita)), aes(label = state), show.legend = FALSE) +
  labs(title = "Yearly HRI Emergency Room Visits per Capita by State",
       x     = "Year",
       y     = "Emergency Room Visits per Capita (100,000)") +
  scale_x_discrete(guide = guide_axis(angle = 90))

summary(lm(emer.per.capita ~ year, data = hri.emer))
```

## Hospitalizations
The first graph plots each observation in the hospitalizations dataset, by the number of visits per capita and the year the observation was recorded. Each point on the plot represents a different state's record. Also plotted in this graph is the least squares regression line, and an imposed line graph that tracks the average hospitalizations per capita, nationwide. The second graph shows the variability between years in this dataset. The third graph highlights the outliers in the second graph.

```{r hosp_cor}
ggplot(hri.hosp, aes(year, hosp.per.capita)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  geom_line(aes(year, hosp.per.capita), data = aggregate(hosp.per.capita ~ year, hri.hosp, mean)) +
  labs(title = "Yearly HRI Hospitalizations per Capita by State",
       x     = "Year",
       y     = "Hospitalizations per Capita (100,000)")

ggplot(hri.hosp, aes(factor(year), hosp.per.capita)) +
  geom_boxplot() +
  labs(title = "Yearly HRI Hospitalizations per Capita by State",
       x     = "Year",
       y     = "Hospitalizations per Capita (100,000)") +
  scale_x_discrete(guide = guide_axis(angle = 90))

ggplot(hri.hosp, aes(factor(year), hosp.per.capita)) +
  geom_boxplot(outlier.colour = "red") +
  geom_text_repel(data = hri.hosp %>% group_by(year) %>% filter(hosp.per.capita > quantile(hosp.per.capita, 0.75) + 1.5*IQR(hosp.per.capita)), aes(label = state), show.legend = FALSE) +
  labs(title = "Yearly HRI Hospitalizations per Capita by State",
       x     = "Year",
       y     = "Hospitalizations per Capita (100,000)") +
  scale_x_discrete(guide = guide_axis(angle = 90))

summary(lm(hosp.per.capita ~ year, data = hri.hosp))
```

## Mortality
Here, the mortality dataset is visualized using similar methods as used for the emergency room visits and hospitalizations datasets. Due to the high variability of the dataset, seen by the large number of outliers, this dataset will not be used for further analysis.

```{r mort_cor}
ggplot(hri.mort, aes(year, mortalities.total)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Yearly HRI Mortalities by State",
       x     = "Year",
       y     = "HRI Related Mortalities (Total)")

ggplot(hri.mort, aes(factor(year), mortalities.total)) +
  geom_boxplot() +
  labs(title = "Yearly HRI Mortalities by State",
       x     = "Year",
       y     = "HRI Related Mortalities (Total)") +
  scale_x_discrete(guide = guide_axis(angle = 90))

ggplot(hri.mort, aes(factor(year), mortalities.total)) +
  geom_boxplot(outlier.colour = "red") +
  geom_text_repel(data = hri.mort %>% group_by(year) %>% filter(mortalities.total > quantile(mortalities.total, 0.75) + 1.5*IQR(mortalities.total)), aes(label = state), show.legend = FALSE) +
  labs(title = "Yearly HRI Mortalities by State",
       x     = "Year",
       y     = "HRI Related Mortalities (Total)") +
  scale_x_discrete(guide = guide_axis(angle = 90))

summary(lm(mortalities.total ~ year, data=hri.mort))
```

## Temperature
Here, we visualize the data in the temperature dataset. The first graph shows the change over time of yearly average temperature in each of the four regions of the United States. The second graph shows the highest three and lowest three outliers by average temperature. The final graph shows the variability in the data across each year.

```{r temp_cor}
ggplot(temp.state, aes(year, temp, fill = state, colour = state.region)) +
        geom_line(show.legend = TRUE) +
        labs(title = "Yearly Average Temperature by State",
             x     = "Year",
             y     = "Mean Yearly Temperature (deg F)",
             colour = "State Region") +
        guides(fill = "none")

ggplot(temp.state, aes(year, temp, fill = state, colour = state)) +
  geom_line(show.legend = FALSE) +
  labs(title = "Yearly Average Temperature by State",
       x     = "Year",
       y     = "Mean Yearly Temperature (deg F)") +
  gghighlight(max(temp) > 67 | min(temp) < 40)


ggplot(temp.state, aes(factor(year), temp)) +
        geom_boxplot() +
        labs(title = "Yearly Average Temperature",
             x     = "Year",
             y     = "Mean Yearly Temperature (deg F)") +
  scale_x_discrete(guide = guide_axis(angle = 90))

summary(lm(temp ~ year, data = temp.state))
```

## Carbon Dioxide
The below graphs explore the atmospheric carbon dioxide emissions dataset. The first graph displays the entirety of the dataset, from 1958 to the present. It also shows the least squares regression line. The second graph displays the same information, but only for the range of years included in this study, from 2000 to 2020.

```{r co2_cor}
ggplot(co2, aes(date, ppm)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Monthly Atmospheric Carbon Dioxide Concentrations (1958 - 2023)",
       x     = "Date",
       y     = "Average Global Atmospheric CO2 Concentration (ppm)")

co2 <- co2 %>% filter(date >= as.Date("2000-01-01"))

ggplot(co2, aes(date, ppm)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Monthly Atmospheric Carbon Dioxide Concentrations (2000 - 2023)",
       x     = "Date",
       y     = "Average Global Atmospheric CO2 Concentration (ppm)")

summary(lm(ppm ~ date, data= co2))
```

# Simple Linear Regressions

This first section correlates each HRI incidence dataset with CO2 concentrations. First the yearly mean CO2 concentration must be calculated, and the national mean each year must also be calculated. Shown is the graph of each dataset with lines of best fit, and the summary of each linear regression model produced.

```{r hri-by-co2}
co2.yearly.mean <- co2 %>%
  group_by(year) %>%
  summarise(ppm = mean(ppm))

hri.emer.yearly.mean <- hri.emer %>%
  group_by(year) %>%
  summarise(emer.per.capita = mean(emer.per.capita))
hri.hosp.yearly.mean <- hri.hosp %>%
  group_by(year) %>%
  summarise(hosp.per.capita = mean(hosp.per.capita))

hri.emer.co2 <- hri.emer.yearly.mean %>%
  left_join(co2.yearly.mean, by = join_by(year))
hri.hosp.co2 <- hri.hosp.yearly.mean %>%
  left_join(co2.yearly.mean, by = join_by(year))

summary(lm(emer.per.capita ~ ppm, data=hri.emer.co2))
summary(lm(hosp.per.capita ~ ppm, data=hri.hosp.co2))

ggplot(hri.emer.co2, aes(ppm, emer.per.capita)) +
  geom_point() +
  geom_smooth(method="lm", se=TRUE) +
  labs(title = "Average National HRI Emergency Room Visits by Yearly Average Atmospheric CO2 Concentration",
       x     = "Global Yearly Average Atmospheric CO2 Concentration (ppm)",
       y     = "National US Emergency Room Visits per Capita (100,000) for HRI")
ggplot(hri.hosp.co2, aes(ppm, hosp.per.capita)) +
  geom_point() +
  geom_smooth(method="lm", se=TRUE) +
  labs(title = "Average National HRI Hospitalizations by Yearly Average Atmospheric CO2 Concentration",
       x     = "Global Yearly Average Atmospheric CO2 Concentration (ppm)",
       y     = "National US Hospitalizations per Capita (100,000) for HRI")
```

Now, each HRI incidence dataset is correlated with state yearly average temperature. The temperature dataset and HRI incidence datasets must be joined by both state and year, and both are required to identify a unique observation. Displayed is the grpah of each joined dataset, and the summaries of each simple linear regression model.

```{r hri-by-temp}
hri.emer.temp <- hri.emer %>%
  left_join(temp.state, by = join_by(state, year)) %>%
  na.omit()
hri.hosp.temp <- hri.hosp %>%
  left_join(temp.state, by = join_by(state, year)) %>%
  na.omit()

summary(lm(emer.per.capita ~ temp, data=hri.emer.temp))
summary(lm(hosp.per.capita ~ temp, data=hri.hosp.temp))

ggplot(hri.emer.temp, aes(temp, emer.per.capita)) +
  geom_point() +
  geom_smooth(method="lm", se=TRUE) +
  labs(title = "HRI Emergency Room Visits by State Yearly Mean Temperature",
       x     = "State Yearly Mean Temperature (deg F)",
       y     = "Emergency Room Visits per Capita (100,000) for HRI")
ggplot(hri.hosp.temp, aes(temp, hosp.per.capita)) +
  geom_point() +
  geom_smooth(method="lm", se=TRUE) +
  labs(title = "HRI Hospitalizations by State Yearly Mean Temperature",
       x     = "State Yearly Mean Temperature (deg F)",
       y     = "Hospitalizations per Capita (100,000) for HRI")
```

# Multiple Linear Regressions

Now, each HRI incidence dataset is correlated with both the indicators in the simple linear regression section. Shown is the summary printouts of each multiple linear regression model, one for emergency room visits and one for hospitalizations.

```{r hri-by-co2-temp}
hri.emer.temp.co2 <- hri.emer.temp %>%
  left_join(co2.yearly.mean, by = join_by(year))
hri.hosp.temp.co2 <- hri.hosp.temp %>%
  left_join(co2.yearly.mean, by = join_by(year))

summary(lm(emer.per.capita ~ ppm + temp, data = hri.emer.temp.co2))
summary(lm(hosp.per.capita ~ ppm + temp, data = hri.hosp.temp.co2))
```

# Model Testing

Here, the data is split into training and testing data. The multiple linear regression model is built upon the training data, and its accuracy in predicting the testing data is measured. The below is for the emergency room visits dataset. Shown is the summary of the fitted linear regression model, then the R-squared of the model when predicting the testing set, then the root mean squared error of the model when predicting the testing set.

```{r hri-by-co2-temp-training-emer}
hri.emer.model <- {}
hri.emer.model$split    <- initial_split(hri.emer.temp.co2)
hri.emer.model$training <- training(hri.emer.model$split)
hri.emer.model$testing  <- testing(hri.emer.model$split)
hri.emer.model$model    <- linear_reg()
hri.emer.model$recipe   <- recipe(emer.per.capita ~ ., data = hri.emer.model$training) %>%
                             update_role(state, new_role = "ID") %>%
                             step_dummy(state.region) %>%
                             step_rm(year, state.id, state.code) %>%
                             step_zv(all_predictors())
hri.emer.model$workflow <- workflow() %>%
                             add_model(hri.emer.model$model) %>%
                             add_recipe(hri.emer.model$recipe)
hri.emer.model$fit <- hri.emer.model$workflow %>% fit(hri.emer.model$training)
summary(hri.emer.model$fit$fit$fit$fit)
hri.emer.model$predict <- predict(hri.emer.model$fit, hri.emer.model$testing) %>%
                            bind_cols(hri.emer.model$testing %>% select(emer.per.capita, state, year))
rsq(hri.emer.model$predict, truth = emer.per.capita, estimate = .pred)
rmse(hri.emer.model$predict, truth = emer.per.capita, estimate = .pred)
```

We repeat the same steps on the hospitalizations dataset.

```{r hri-by-co2-temp-training-hosp}
hri.hosp.model <- {}
hri.hosp.model$split    <- initial_split(hri.hosp.temp.co2)
hri.hosp.model$training <- training(hri.hosp.model$split)
hri.hosp.model$testing  <- testing(hri.hosp.model$split)
hri.hosp.model$model    <- linear_reg()
hri.hosp.model$recipe   <- recipe(hosp.per.capita ~ ., data = hri.hosp.model$training) %>%
                             update_role(state, new_role = "ID") %>%
                             step_dummy(state.region) %>%
                             step_rm(year, state.id, state.code) %>%
                             step_zv(all_predictors())
hri.hosp.model$workflow <- workflow() %>%
                             add_model(hri.hosp.model$model) %>%
                             add_recipe(hri.hosp.model$recipe)
hri.hosp.model$fit <- hri.hosp.model$workflow %>% fit(hri.hosp.model$training)
summary(hri.hosp.model$fit$fit$fit$fit)
hri.hosp.model$predict <- predict(hri.hosp.model$fit, hri.hosp.model$testing) %>%
                            bind_cols(hri.hosp.model$testing %>% select(hosp.per.capita, state, year))
rsq(hri.hosp.model$predict, truth = hosp.per.capita, estimate = .pred)
rmse(hri.hosp.model$predict, truth = hosp.per.capita, estimate = .pred)
```